---
output:
    pdf_document:
        latex_engine: xelatex
        template: './pandoc-templates/template.tex'
        keep_tex: false
bibliography: './bib/library.bib'
csl: './bib/vancouver.csl'
---

```{r setup, include = FALSE}
library(ggplot2)
library(scales)
library(gridExtra)
library(cowplot)
library(pander)
library(xtable)
    options(xtable.caption.placement = 'top',
            xtable.include.rownames = FALSE,
            xtable.booktabs = TRUE,
            xtable.comment = FALSE)
library(dplyr)
library(tidyr)
library(knitr)
library(readr)
library(stringr)
    
opts_chunk$set(echo = FALSE,
               warning = FALSE,
               message = FALSE,
               cache = FALSE,
               fig.align = 'center',
               fig.cap = 'character',
               dev = 'pdf',
               tidy = FALSE)

```

# General items

## Summary statement of the proposal for inclusion.

We are applying for the _**inclusion of gabapentin as an analgesic agent for the management of neuropathic pain (central and peripheral) in adults**_. The medicine has regulatory approval for the treatment of several neuropathic pain states in adults by numerous stringent regulatory bodies (including the Food and Drug Administration [@NeurontinFDA2015] and European Medicines Agency [@NeurontinEMA2013]). Furthermore, all recent evidence-based treatment guidelines recommend gabapentin as one of the first-line agents for the pharmacological management of neuropathic pain of central or peripheral origin [@Finnerup2015; @NICE2013; @SIGN2013; @Attal2010].

A recent systematic review estimated the prevalence of neuropathic pain in the general, adult population to be between 7 and 10% [@VanHecke2014], equating to over 518 million prevalent cases of adults with neuropathic pain globally. And, in certain chronic diseases that already impose or are predicted to impose a high burden of disease in developing countries, such as HIV-AIDS, diabetes mellitus, leprosy, and low-back pain, the prevalence of neuropathic pain can be more than three times the population prevalence [@Sadosky2008; @Thakur2015; @Rice2016]. In addition, developing countries are disproportionally affected by acute traumatic injuries (e.g., conflict-related trauma, motor vehicle injuries) that may cause nerve damage [@Gosselin2009]. Having neuropathic pain has a major negative impact on health-related quality of life, and places a significant human and economic burden on health resources [@Doth2010; @Torrance2014].

Neuropathic pain is difficult to treat, and requires specific classes of medication for its management. Evidence-based recommendations include: tricyclic antidepressants (TCAs), $\alpha_{2}\delta$ calcium channel ligands (gabapentin and pregabalin), and serotonin and noradrenaline reuptake inhibitors (SNRIs, duloxetine and venlafaxine) as first-line agents. The number needed to treat (NNT) to achieve 50% pain relief relative to placebo for these effective medications ranges between 4 and 9 [amitriptyline: 4.3 (95% CI: 3.6 to 5.3), gabapentin 6.3 (95% CI: 5.0 to 8.3)] [@Finnerup2015; @NICE2013; @Attal2010], and failure to respond adequately to initial monotherapy necessitates switching to another class of agent, or using combination therapy. Thus, management of neuropathic pain requires an adequate armamentarium of medications that have proven efficacy and may be used in combination. 

The WHO recently urged member states to ensure, _“the availability of essential medicines for the management of symptoms, including pain,”_
and _“[the] education and training of healthcare professionals, in order to ensure adequate responses to palliative care needs.”_ [@WHO2014]. Yet for neuropathic pain, the WHO Model List of Essential Medicines [@WHO2015a; @WHO2015b] is deficient in drugs with proven efficacy in treating neuropathic pain. The Model List includes only one medicine with evidence-based recommendations as a first-line treatment (amitriptyline), and no agents that are recommended as second-line treatments. And, the WHO Model Formulary [@WHO2008] provides out-dated and erroneous guidance on appropriate medications to use for treating neuropathic pain. These deficiencies are echoed in the national essential medicines lists of developing and emerging countries [@Kamerman2015]. 

Given its proven efficacy, good cost-utility, and global availability, we are therefore applying for inclusion of gabapentin as an additional treatment for neuropathic pain on the Model Essential Medicines List.  

\newpage

## Name of the WHO technical department and focal point supporting the application (where relevant).

## Name of organization(s) consulted and/or supporting the application.

#### Proposing organizations

- International Association for the Study of Pain (IASP);

- Neuropathic Pain Special Interest Group (NeuPSIG) of the IASP;

- International Association for Hospice and Palliative Care (IAHPC)

#### Supporting organizations 
_(see Appendix 1 for copies of the letters of support)_

- [to be inserted later]

## International Nonproprietary Name (INN) and Anatomical Therapeutic Chemical (ATC) code of the medicine.

```{r name, results = 'asis'}

# Make dataframe
foo <- data.frame(`Taxonomic system` =
                      c('International Nonproprietary Name (INN)',
                        'Anatomical Therapeutic Chemical (ATC)'),
                  Name = c('Gabapentin', 'N03AX12'))

# Make xtable
name_chunk <- xtable(foo,
                     caption = 'Drug classification',
                     label = 'drug_class',
                     align = c('l', 'l', 'l'),
                     display = c('s', 's', 's'))

# Print table
print(name_chunk)

# Clean environment
rm(list = ls())

```

## Formulation(s) and strength(s) proposed for inclusion; including adult and paediatric (if appropriate).

Gabapentin is only approved for use in managing neuropathic pain in adults. 

#### Core List

- Tablets:  100mg, 200mg, 300mg, 400mg, 600mg, 800mg

- Capsules: 100mg, 300mg, 400mg, 600mg


#### Complementary List

- Oral solution: 50 mg/ml (150-470ml bottle sizes available through different manufacturers)

## International availability
Table \ref{availability} lists countries, trade names, formulations, and manufacturers of gabapentin obtained from Martindale: The Complete Drug Reference [@Martindale2014] via Micromedex Solutions (Micromedex Inc., http://micromedex.com). The source listed 206 unique manufacturers of gabapentin across 42 countries, marketing the medicine under 241 proprietary names. Of these, 17 countries, 161 manufacturers, and 177 proprietary names were listed under countries classified as emerging or developing countries by the International Monetary Fund [@IMF2016]. 

The 17 developing or emerging countries where gabapentin was listed as being available constitute about `r round((43847277 + 209567920 + 18131850 + 1382323332 + 9821318 + 1326801576 + 260581100 + 30751602 + 128632004 + 102250133 + 38593161 + 143439832 + 54978907 + 68146609 + 79622062 + 44624373 + 31518855)/7432663275, 2) * 100 `% of the global population [@UN2015]. 

All listed products were for the tablet and capsule formulations of gabapentin.

```{r availability, results = 'asis'}

############################################################
#                                                          #
#          Availability table: See availability.R          #
#            and availability.RData for details            #
#                                                          #
############################################################

# Load data
load('./data/availability.RData')

# Make xtable
xtab_availability <- xtable(data.merged,
                            caption = 'International availability of gabapentin \\newline \\textit{(emerging and developing nations are highlighted)}',
                            label = 'availability',
                            align = rep('l', 5),
                            display = rep('s', 5))

# Add additional commands to header rows 
addtorow <- list(pos = as.list(c(0, emphasis.index - 1)),
                 command = c(paste0("\\hline \n",
                                    "\\endfirsthead \n",
                                    "\\hline \n",
                                    "{\\footnotesize Continued on next page} \n",
                                    "\\endfoot \n",
                                    "\\bottomrule \n",
                                    "\\endlastfoot \n",sep=""),
                             rep("\\rowcolor[gray]{.9}",
                                 length(emphasis.index))))

# Print table
print(xtab_availability,
      floating = FALSE,
      tabular.environment = "longtable",
      add.to.row = addtorow,
      hline.after = c(-1))

# Clean environment
rm(list = ls())
```

## Whether listing is requested as an individual medicine or as representative of a pharmacological class.

We are requesting the inclusion of gabapentin as an individual medicine.

# Treatment details, public health relevance and evidence appraisal and synthesis

## Treatment details (requirements for diagnosis, treatment and monitoring)

### Diagnosis and monitoring

The diagnosis of neuropathic pain can be established using a history and clinical examination, and without the need for specialised equipment or facilities [@Finnerup2016; @Haanpaa2009; @Haanpaa2011]. Figure \ref{grading} outlines the diagnostic process and how each step informs the level of diagnostic certainty [@Finnerup2016]. Like the diagnosis, monitoring of treatment outcome can be performed without specialised equipment or facilities. Readily available clinical screening tools such as the Douleur Neuropathique en 4 questions (DN4), Leeds Assessment of Neuropathic Symptoms and Signs (LANSS), and painDETECT can be used to assist in diagnosing pain of neuropathic origin. These tools have been translated and validated into numerous languages [@Bennett2007]. 


\begin{figure}[ht]
\centering
\includegraphics[scale = 0.5, angle = -90]{./images/grading}
\captionsetup{justification=justified}
\caption{Diagnosis and grading of neuropathic pain. The level of \textit{"probable"} is usually sufficient to initiate treatment according to neuropathic pain guidelines. The level \textit{"definite"} is useful in specialist contexts and when a causal treatment of the underlying lesion or disease is an option. \textsuperscript{a} History, including pain descriptors, the presence of non-painful sensory symptoms, and aggravating and alleviating factors, suggestive of pain being related to a neurological lesion. The suspected lesion or disease is reported to be associated with neuropathic pain, including a temporal and spatial relationship representative of the condition. \textsuperscript{b} The pain distribution reported by the patient is consistent with the suspected lesion or disease. \textsuperscript{c} The area of sensory changes may extend beyond, be within, or overlap with the area of pain. Sensory loss is generally required but touch-evoked or thermal allodynia may be the only finding at bedside examination. \textsuperscript{d} The term \textit{"definite"} in this context means \textit{"probable neuropathic pain with confirmatory tests"}. Adapted from: Finnerup et al., 2016 [17].}
\label{grading}
\end{figure}

### Treatment

The information on treatment was obtained from regulatory documents available from the Food and Drug Administration (FDA) [@NeurontinFDA2015], and European Medicines Agency (EMA) [@NeurontinEMA2013] for Neurontin (gabapentin, Pfizer Inc).

#### Dosage and administration

Usual dosage range:  

- _Adults:_ 900-1800mg/day in three divided doses.

- _Children:_ Gabapentin is not approved for the management of neuropathic pain in children.

    We are cognisant of the lack of therapeutic choices for children, and that gabapentin is indicated for paediatric use for epilepsy by major regulatory bodies. But, while gabapentin has demonstrative evidence of tolerability and safety in children there is insufficient data on the use of the drug for the treatment of neuropathic pain in children to draw evidence-based recommendations. 
    
    A recent review of neuropathic pain in children provides an excellent summary of our knowledge of neuropathic pain and its treatment in children: _"The most common neuropathic pain conditions seen in adults are rare in children. Some neuropathic conditions are becoming increasingly recognized in children and adolescents, including complex regional pain syndromes, phantom limb pain, spinal cord injury, trauma and postoperative neuropathic pain, autoimmune and degenerative neuropathies (eg, Guillain-Barr$\acute{e}$ syndrome, Charcot-Marie-Tooth disease), and the effects of cancer disease processes and treatment. Some neuropathic pain syndromes are relatively unique to the pediatric population, including toxic and metabolic neuropathies (eg, lead, mercury, alcohol, infection), hereditary neurodegenerative disorders (eg, Fabry disease), mitochondrial disorders, and primary erythromelalgia. All these cause significant suffering to children and their caregivers and steps need to be taken to alleviate this suffering....In some countries, gabapentin has been approved for the use of paediatric neuropathic pain. However, the amount of evidence available on the effectiveness and safety of gabapentin in pediatric neuropathic pain is too weak for the authors to make a recommendation at this time. Additional studies are recommended and needed."_ [@Walco2010].
    
    Unfortunately, the evidence-base for treatments of neuropathic pain in children has not advanced significantly since the writing of the review. But hopefully if recent activity in this area continues [@Brown2016], the evidence for the use of gabapentin for neuropathic pain in children can be reassessed for future editions of the Essential Medicines List for Children. 
    
Gabapentin can be administered with or without food, and should be swallowed whole with sufficient fluid (e.g. a glass of water).

The FDA and the EMA recommend the following titration schedule for neuropathic pain:

- Day 1: Single 300mg dose

- Day 2: 600mg/day (300mg two times a day)

- Day 3: 900mg/day (300mg three times a day)

Thereafter, based on therapeutic response and tolerability, the dose can be increased in 300mg/day increments every 2 to 3 days up to a maximum dose of 1800mg/day. Slower titration of gabapentin dosage may be appropriate for individual patients. The minimum time to reach a dose of 1800 mg/day is one week. In analgesic clinical trials of gabapentin in neuropathic pain, the clinical effect (separation from placebo) typically was evident by the end of first week of treatment.

In clinical studies, efficacy was demonstrated over a range of doses from 1800mg/day to 3600mg/day with comparable effects across the dose range. However, the additional benefit of using doses greater than 1800mg/day was not demonstrated. If a patient is to be prescribed doses greater than 1800mg/day, the EMA recommend a minimum titration time of two weeks to reach 2400mg/day, and three weeks to reach 3600mg/day. 

If the gabapentin dose is reduced, discontinued, or substituted with an alternative medication, this should be done gradually over a minimum of one week (a longer period may be needed at the discretion of the prescriber).

#### Special populations

**Children:** Gabapentin does not have regulatory approval for managing neuropathic pain in paediatric patients, but does have approval for use as adjunctive therapy or mono-therapy for partial seizures in children aged 3 to 11 years. Dosing recommendations for this indication are not presented here.

**Elderly:** The total number of patients treated with gabapentin in clinical trials in patients with post-herpetic neuralgia was 336, of which 102 (30%) were 65 to 74 years of age, and 168 (50%) were 75 years of age and older. There was a larger treatment effect in patients 75 years of age and older compared with younger patients who received the same dosage. Since gabapentin is almost exclusively eliminated by renal excretion, the larger treatment effect observed in patients 75 years or older may be a consequence of increased gabapentin exposure for a given dose that results from an age-related decrease in renal function. However, other factors cannot be excluded. Nevertheless, care should be taken in dose selection, in the elderly, and dose should be adjusted based on creatinine clearance values in these patients. The types and incidence of adverse reactions were similar across age groups except for peripheral oedema and ataxia, which tended to increase in incidence with age.

**Renal impairment:** Gabapentin undergoes almost exclusive renal clearance, and therefore it is recommended that the dose of gabapentin is reduced in patients with renal impairment or undergoing haemodialysis (paediatric patients with renal insufficiency have not been studied). Recommended dosage adjustments are shown in Table \ref{renal}.

\begin{table}[h!]
% Had to make table by hand because of complicated multi-column format
\centering
\caption{Gabapentin dosage based on renal function}
\label{renal}
\begin{tabular}{@{}lllllll@{}}
\toprule
\begin{tabular}[b]{@{}l@{}}Renal function: \\ creatinine \\ clearance \\ (ml/min)$^{\ast}$\end{tabular} & \begin{tabular}[b]{@{}l@{}}Total daily \\ dose range \\ (mg/day)\end{tabular} & \begin{tabular}[b]{@{}l@{}}Regimen:$^{\dag}$ \\ A\end{tabular} & B           & C           & D           & E            \\ \toprule
$\geq$ 60                                                                                      & 900 to 3600                                                                   & 300$_{tid}$                                          & 400$_{tid}$ & 600$_{tid}$ & 800$_{tid}$ & 1200$_{tid}$ \\
30 to 59                                                                                       & 400 to 1400                                                                   & 200$_{bid}$                                          & 300$_{bid}$ & 400$_{bid}$ & 500$_{bid}$ & 700$_{bid}$  \\
15 to 29                                                                                       & 200 to 700                                                                    & 200$_{qd}$                                           & 300$_{qd}$  & 400$_{qd}$  & 500$_{qd}$  & 700$_{qd}$   \\
$<$ 15$^{\ddag}$                                                                            & 100 to 300                                                                    & 100$_{qd}$                                           & 125$_{qd}$  & 150$_{qd}$  & 200$_{qd}$  & 300$_{qd}$   \\ 
\bottomrule
\multicolumn{7}{l}{} \\
\multicolumn{2}{l}{}                                                                                                                                                           & \multicolumn{5}{l}{Post-haemodialysis supplemental dose (mg)$^{\mathbb{S}}$}                  \\
\toprule
\multicolumn{2}{l}{Haemodialysis}                                                                                                                                               & 125                                                  & 150         & 200         & 250         & 350          \\ \bottomrule
\multicolumn{7}{L{130mm}}{\footnotesize{\textbf{$\dag:$} $tid$ = three times a day, $bid$ = two times a day, $qd$ = single daily dose }} \\
\multicolumn{7}{L{130mm}}{\footnotesize{\textbf{$\ddag:$} For patients with creatinine clearance $<$ 15ml/min, reduce daily dose in proportion to creatinine clearance (e.g., patients with a creatinine clearance of 7.5ml/min should receive one-half the daily dose that patients with a creatinine clearance of 15ml/min receive).}} \\
\multicolumn{7}{L{130mm}}{\footnotesize{\textbf{$\mathbb{S}:$} Patients on haemodialysis should receive maintenance doses based on estimates of creatinine clearance as indicated in the upper portion of the table and a supplemental post-haemodialysis dose administered after each 4 hours of haemodialysis as indicated in the lower portion of the table.}} \\
\multicolumn{7}{L{130mm}}{\footnotesize{\textbf{$\ast:$} In out-patients with stable renal function, creatinine clearance (CLCr) can be reasonably well estimated using the equation of Cockcroft and Gault: \newline
$CLCr = \dfrac{[140~-~age~(years)]~\times~weight~(kg)}{72~\times~serum~creatinine~(mg/dl)}~(\times~0.85~for~female~patients)$}}\\
\end{tabular}
\end{table}

**Pregnancy:** Gabapentin is classified as _Pregnancy Category C_. There are no adequate and well-controlled studies in pregnant women. In pre-clinical studies in mice, rats, and rabbits, gabapentin was developmentally toxic when administered to pregnant animals at doses similar to or lower than those used clinically. Gabapentin should be used during pregnancy only if the potential benefit justifies the potential risk to the fetus.

**Nursing Mothers:** Gabapentin is secreted into human milk following oral administration, possibly exposing a nursed infant to a maximum dose of approximately 1 mg/kg/day of gabapentin. Because the effect on the nursing infant is unknown, gabapentin should be used in women who are nursing only if the benefits clearly outweigh the risks.

**Other:** Although no formal studies have been conducted, neither sex, race nor hepatic impairment have been reported to affect the pharmacokinetics of gabapentin.

#### Pharmacokinetics

**Oral Bioavailability:** Gabapentin bio-availability is not dose proportional. That is, as dose is increased, bio-availability decreases. Bio-availability of gabapentin is approximately 60%, 47%, 34%, 33%, and 27% following 900, 1200, 2400, 3600, and 4800 mg/day given in 3 divided doses, respectively. Food has only a slight effect on the rate and extent of absorption of gabapentin [14% increase in area under the curve (AUC) and maximum plasma concentration (Cmax)].

**Distribution:** Less than 3% of gabapentin circulates bound to plasma protein. The apparent volume of distribution of gabapentin after 150mg intravenous administration is 58 $\pm$ 6 l (mean $\pm$ SD). In patients with epilepsy, the minimum plasma concentrations (Cmin) of gabapentin in cerebrospinal fluid were approximately 20% of the corresponding plasma concentrations.

**Elimination:** All pharmacological actions following gabapentin administration are due to the activity of the parent compound. Gabapentin is not appreciably metabolised in humans, and is eliminated from the systemic circulation by renal excretion as unchanged drug. As such, gabapentin elimination rate constant, plasma clearance, and renal clearance are directly proportional to creatinine clearance. Gabapentin elimination half-life is 5 to 7 hours and is unaltered by dose or following multiple dosing.

In elderly patients, and in patients with impaired renal function, gabapentin plasma clearance is reduced. If required, gabapentin can be removed from plasma by haemodialysis.

**Long-term use and overdose:** The EMA documentation notes that the efficacy and safety of gabapentin has not been examined in clinical studies for treatment periods longer than five months, and if a patient requires dosing for longer periods that the treating physician should assess the patient’s clinical status and determine the need for additional therapy.

A lethal dose of gabapentin was not identified in mice and rats receiving single oral doses as high as 8000mg/kg. Signs of acute toxicity in animals included ataxia, laboured breathing, ptosis, sedation, hypoactivity, or excitation. Acute oral overdoses of gabapentin up to 49000mg have been reported. In these cases, double vision, slurred speech, drowsiness, lethargy, and diarrhoea were observed. All patients recovered with supportive care. Coma, resolving with dialysis, has been reported in patients with chronic renal failure who were treated with gabapentin.

#### Drug interactions

**_In vitro_ studies:** _In vitro_ studies investigating the effect of gabapentin on major cytochrome P450 enzymes (CYP1A2, CYP2A6, CYP2C9, CYP2C19, CYP2D6, CYP2E1, and CYP3A4) found that the only isoform inhibited was CYP2A6, and only at the highest dose tested (171$\mu$g/ml, ~15 times the Cmax at 3600mg/day). Even then, the level of inhibition was modest (14 to 30%).

**_In vivo_ studies:** _In vivo_ drug interaction studies for gabapentin have involved healthy adults and adult patients with epilepsy. The following medicines displayed no interaction with gabapentin:

- Phenytoin

- Carbamazepine

- Valproic Acid

- Phenobarbital

- Naproxen _(Doses of gabapentin and naproxen used were lower than the therapeutic doses for both drugs. The magnitude of interaction within the recommended dose ranges of either drug is not known.)_

- Probenecid _(Probenecid is a blocker of renal tubular secretion, but did not alter gabapentin pharmacokinetic parameters, indicating that gabapentin does not undergo renal tubular secretion by the pathway that is blocked by probenecid.)_ 

Of importance, there is no known interactions between gabapentin and agents recommended as first-line treatments HIV infection in adults, breast-feeding women, adolescents, and children (lamivudine, abacavir, zidovudine, tenofovir, stavudine, lopinavir/ritonavir, darunavir, dolutegravir, efavirenz, emtricitabine, nevirapine) [@WHO2016]. Nor does gabapentin interact with first-line treatments for tuberculosis infection (isoniazid, rifampicin, streptomycin, ethambutol, pyrazinamide) [@WHO2010], or treatment of diabetes mellitus (insulin, metformin, sulfonylureas) [@WHO2014]. 

Interactions have been shown with the following medicines:

- Hydrocodone: Co-administration of gabapentin (125 to 500mg) decreased hydrocodone (10 mg) Cmax and AUC values in a dose-dependent manner. Cmax and AUC values were 3% to 4% lower, respectively, after administration of 125mg gabapentin, and 21% to 22% lower, respectively, after administration of 500mg gabapentin. Hydrocodone increased gabapentin AUC values by 14%. The magnitude of the interaction at other doses is not known.
    
- Morphine: A literature article reported that when a 60mg controlled-release morphine capsule was administered 2 hours before 600mg of gabapentin, mean gabapentin AUC increased by 44% compared to gabapentin administered without morphine. Morphine pharmacokinetic parameters were not affected by administration gabapentin. The magnitude of the interaction at other doses is not known.

- Cimetidine: Administration of cimetidine (300mg, three times daily), was associated with a 14% reduction in the clearance of gabapentin. Creatinine clearance fell by 10%, indicating that cimetidine may alter the renal excretion of both gabapentin and creatinine. This small decrease in excretion of gabapentin by cimetidine is not expected to be of clinical importance. The effect of gabapentin on cimetidine was not evaluated.

- Oral contraceptive: The AUC and half-life of multiple-dose pharmacokinetic profiles of norethindrone (50$\mu$g) and ethinyl estradiol (2.5mg) were similar with and without co-administration of gabapentin (400mg three times a day). However, the Cmax of norethindrone was 13% higher when co-administered with gabapentin. This interaction is not expected to be of clinical importance.

- Antacid: Administration of antacid containing aluminium hydroxide and magnesium hydroxide reduced the mean bio-availability of gabapentin by about 20%. This decrease in bio-availability was about 10% when gabapentin was administered 2 hours after the antacid.

#### Drug abuse and dependence

Gabapentin is not a scheduled drug.

The dependence and abuse potential of gabapentin has not been formally evaluated in human studies.

**Abuse:** Gabapentin does not exhibit affinity for benzodiazepine (GABA), opioid, or cannabinoid-1 receptor sites. A small number of post-marketing cases report gabapentin misuse and abuse. These individuals were taking higher than recommended doses of gabapentin for unapproved uses. Most of the individuals described in these reports had a history of poly-substance abuse or used gabapentin to relieve symptoms of withdrawal from other substances. Consequently, the FDA recommends that patients with a history of drug abuse should be carefully evaluated before starting gabapentin. And, they should be observed for signs and symptoms of gabapentin misuse or abuse (e.g., development of tolerance, self-dose escalation, and drug-seeking behaviour). The FDA recommendation is consistent with a review of the literature by Schifano [@Schifano2014], which concluded that the risk of misuse of $\alpha_{2}\delta$ calcium channel ligands is low when the drugs are administered at therapeutic doses to individuals with no history of substance misuse.

**Dependence:** 
There are rare post-marketing reports of individuals experiencing withdrawal symptoms shortly after discontinuing higher than recommended doses of gabapentin used to treat illnesses for which the drug is not approved. Such symptoms included agitation, disorientation and confusion after suddenly discontinuing gabapentin. These symptoms resolved after restarting gabapentin. Most of these individuals had a history of poly-substance abuse or used gabapentin to relieve symptoms of withdrawal from other substances.

Increased seizure frequency may occur in patients with seizure disorders if gabapentin is abruptly discontinued  

### Guideline recommendations

We are unaware of any WHO guidelines for the treatment of neuropathic pain, but several reputable organizations that are independent of the WHO have developed evidence-based guidelines. These include:

- Pharmacotherapy for Neuropathic Pain in Adults: A Systematic Review and Meta-analysis, Association for the Study of Pain (IASP) [@Finnerup2015] _(Please note that N Finnerup, S Haroutonioun, PR Kamerman, SN Raja, ASC Rice and BH Smith were involved in the development of this guideline)_;

- Neuropathic Pain: The Pharmacological Management of Neuropathic Pain in Adults in Non-Specialist Settings, National Institute for Health and Care Excellence (NICE), UK [@NICE2013];

- EFNS Guidelines on the Pharmacological Treatment of Neuropathic Pain: 2010 Revision, European Federation of Neurological Societies [@Attal2010].

All three guidelines agree that tricyclic antidepressants, $\alpha_{2}\delta$ calcium channel ligands (gapapentin and pregabalin), and selective serotonin and noradrenaline re-uptake inhibitors should be considered first-line therapy. With the choice of medicine being guided by clinical and therapeutic factors (e.g., contraindications, drug interactions), and medicine availability and affordability.  

## Information supporting the public health relevance
Neuropathic pain is defined as _"Pain caused by a lesion or disease of the somatosensory nervous system"_ [@Jensen2011; @Treede2008]. It is commonly associated with back pain (e.g., lumbar or cervical radiculopathy), diabetes (painful diabetic neuropathy), post-surgical pain, HIV-AIDS, and herpes zoster (post-herpetic neuralgia), but can also arise through many other diseases or injuries. Specific clinical features include symptoms such as paraesthesia, burning or shooting pains, altered sensation (numbness, allodynia or hyperalgesia), and locally altered autonomic function [@McMahon2013].

In the absence of a 'gold standard' for defining cases and a clinical code for routine healthcare use, it is impossible to identify the precise prevalence of neuropathic pain, for example through the Global Burden of Disease 2013 study [@Global2015]. However, a recent systematic review found that between 7 and 10% of the adult population are affected by pain with neuropathic characteristics (identified through validated questionnaires) [@VanHecke2014]. With a global population of approximately 7.4 billion people, this means that some 518 to 740 million individuals are estimated to currently be affected by neuropathic pain. This includes (but is not restricted to) people with:

- diabetes mellitus (410 million prevalent cases globally, increasing by 133% since 1990 [@Global2015], and projected to rise further [@NCD2016]). Approximately 26% of those with diabetes mellitus have painful polyneuropathy [@VanHecke2014; @Davies2006], equating to 107 million individuals. Figure \ref{diabetes} shows estimates and projected prevalence data for diabetes mellitus from 1985 to 2030 [@NCD2016; @Guariguata2014], with the estimated number of coincident cases of painful polyneuropathy [@Veves2008]. 


```{r diabetes, eval=FALSE}

############################################################
#                                                          #
#       Diabetes figure: See diabetes.R for details        #
#                                                          #
############################################################

```

\begin{figure}
\centering
\includegraphics[scale = 0.5]{./images/diabetes}
\captionsetup{justification=justified}
\caption{Estimated and projected number of cases of diabetes mellitus (blue) between 1985 and 2035, and the number of cases of painful diabetic polyneuropathy (orange) over the same time period based on conservative estimates of between 10 and 20\% of individuals with diabetes developing a painful neuropathy.}
\label{diabetes}
\end{figure}

- HIV/AIDS (29 million prevalent cases globally, increasing by 275% since 1990 [@Global2015]). Approximately 35% of people with HIV/AIDS have painful neuropathy [@Sadosky2008], equating to 10 million individuals. The incidence [@Mehta2011; @Hung2008] and prevalence [@Ellis2010] of the neuropathy has decreased since the introduction of newer antiretroviral regimens that forego neurotoxic medicines such as stavudine, but remains high [@Kamerman2016b].

- chronic low back pain (651 million prevalent cases globally, increasing by 57% since 1990 [@Global2015]). Approximately 37% of those with chronic low back pain have been shown to experience neuropathic pain [@Freynhagen2006], equating to 228 million individuals;

Trauma also a major cause of nervous system injury, and hence neuropathic pain. Data from the Global Burden of Disease initiative indicate that physical injury is more common in developing countries than in developed countries [@Global2015], and thus those with the least resources are likely to face a greater burden of trauma-related neuropathic pain. This greater burden is superimposed on the already greater risk for neuropathic pain in these regions associated with increasing prevalence of diabetes, and a disproportionate share of conditions such as HIV/AIDS and leprosy (Figure \ref{ghdx}).

```{r ghdx, eval = FALSE}

############################################################
#                                                          #
#                 Global burden of health                  #
#              figure: See ghdx.Rfor details               #
#                                                          #
############################################################

```

\begin{figure}
\centering
\includegraphics[scale = 0.5]{./images/ghdx}
\captionsetup{justification=justified}
\caption{Disability Adjusted Life Years (DALY) in developed (orange), and developing (blue) countries associated with diabetes mellitus, HIV/AIDS, leprosy, road injuries, interpersonal violence, and forces of nature, war, and legal intervention for the period 1990 and 2013. The shaded areas show uncertainty estimates. Data were downloaded from \href{http://ghdx.healthdata.org}{Global Health Data Exchange: GHDx}  on 20 July 2016.}
\label{ghdx}
\end{figure}

Older age is one of the most important risk factors for neuropathic pain [@Smith2011]. The ageing population worldwide, as well as the separate rising prevalence of underlying conditions such as diabetes mellitus [@Global2015; @NCD2016] mean that neuropathic pain is likely to increase in prevalence and importance in the future.

Neuropathic pain has a significant adverse impact on all measured aspects of life, health and function [@Torrance2006]. This impact is greater than the impact of chronic, non-neuropathic pain, even when adjusting for pain intensity [@Smith2007], and is irrespective of the underlying diagnosis [@Doth2010]. In one study, 17% of people reporting neuropathic pain rated their quality of life as ‘worse than death’, according to the validated EQ5D measure [@Torrance2014]. Average quality of life scores in the presence of neuropathic pain are comparable to those in severe depression, in poorly-controlled DM, and after recent myocardial infarction [@Smith2007].

In general, neuropathic pain responds poorly to treatment with conventional analgesics (there is no evidence for effectiveness of medicines such as non-steroidal anti-inflammatory drugs [@Moore2015]), and specific classes of medication are required. Gabapentin is recommended as a first-line treatment for neuropathic pain in many national and international guidelines [@Finnerup2015; @NICE2013; @SIGN2013; @Attal2010; @Moulin2007]. Tricyclic antidepressants (TCAs) are also recommended first-line treatments in these guidelines, and are already widely available, and cheaper than gabapentin [@Kamerman2015]. The target population for gabapentin use is therefore all those with neuropathic pain who have not responded, or not responded sufficiently to TCAs, or for whom TCAs are contra-indicated (e.g., glaucoma, cardiovascular disorders, epileptic seizures, poly-pharmacy) or not tolerated. The target population excludes those in whom gabapentin is contraindicated (e.g., in renal failure) or have a known intolerance to gabapentin.

The effectiveness of medicines used in neuropathic pain was recently reviewed systematically and comprehensively [@Finnerup2015]. In this study, gabapentin (excluding extended release preparations and the pro-drug enacarbil) had a demonstrated number needed to treat (NNT) of 6.3 to achieve at least 50% reduction in pain severity scores relative to placebo. If, as above, 518 million people have neuropathic pain worldwide, the use of gabapentin will potentially lead to this successful treatment outcome for around 82 million people. Excluding the 144 million people who could potentially achieve 50% reduction in pain from TCAs (with an NNT of 3.6 [@Finnerup2015], there remain approximately 59 million individuals who could potentially achieve this outcome from gabapentin.  

The actual number who could benefit will be higher because (a) many more will achieve important reductions in pain severity, though less than 50%; (b) TCAs are contraindicated in many people – for example, they are not recommended for use in older adults [@Beer2015]; (c) combination of gabapentin and TCA are effective and recommended [@NICE2013; @SIGN2013]; and (d) these NNTs are calculated after adjusting for placebo and other non-specific effects, so the actual effectiveness is greater than they suggest.

Our review of national essential medicine lists for medications recommended as first- or second-line treatments for neuropathic pain identified that the majority of countries reviewed only had one class of first-line treatment listed (typically a TCA), and about 40% had no second-line treatments listed (Figure \ref{eml}. Of the countries listing two or more first-line medications, the most commonly listed agent was gabapentin (30% of all countries) [@Kamerman2015]. Most of the countries did, however, list morphine (95%), a medication on the Model List with evidence supporting its use in the treatment of neuropathic pain. But, the evidence supporting the use of morphine and other strong opioids in neuropathic pain is of low quality [@McNicol2013; @Attal2010], and this information, together with questions about the safety of strong opioids (e.g., high rates of adverse effects and study withdrawal due to adverse effects, and dependency concerns) means that strong opioids typically are recommended as third-line treatments for neuropathic pain [@Finnerup2015; @Attal2010; @NICE2013]. Thus, the majority of the 104 developing and emerging countries' essential medicine lists had a very limited scope of first- and second-line treatments for neuropathic pain. This limitation is counter to WHO Resolution EB134.R7 of 2014 [@WHO2014], which urges member states to ensure, _“the availability of essential medicines for the management of symptoms, including pain,”_. 

\begin{figure}[h!]
\centering
\includegraphics[scale = 0.6, angle = -90]{./images/eml}
\captionsetup{justification=justified}
\caption{Percentage of national essential medicine lists (NEMLs) that included 0, 1, 2, or 3 drug classes recommended for the treatment of neuropathic pain. Data are shown grouped according to World Bank income category and for all countries (n = 109, data from the Cook Islands, Nauru, and Niue were not included because the World Bank does not index them). The top panel shows drug-classes recommended as first-line treatment, and the bottom panel shows second-line drug classes. First-line drug classes include: tricyclic antidepressants, serotonin and norepinephrine reuptake inhibitors, and $\alpha 2 \delta$ calcium channel ligands. Second-line drug classes include: opioids (including tramadol) and topical agents (5\% lidocaine). There was a positive association between income category and the number of first-line and second-line drug classes listed on NEMLs (corrected P $<$ 0.001). $\ast$ The tricyclic antidepressant amitriptyline was the only first-line drug listed on the NEMLs of 32\% of low-income countries, 36\% of lower-middle income countries, 28\% of higher-middle income countries and 4\% of high-income countries. From: Kamerman et al., 2015 [13].}
\label{eml}
\end{figure}

## Review of benefits: summary of comparative effectiveness in a variety of clinical settings.

The treatment of neuropathic pain is pharmacologically based as there is scant evidence from high-quality placebo-controlled trials supporting the use of invasive procedures [@Dworkin2013] or psychological or behaviour-based therapies.

For pharmacological interventions the evidence supporting this application is based upon our recent systematic review, meta-analysis and GRADE based clinical guideline formulation [@Finnerup2015]. The systematic review of the literature used a standardised review and data extraction protocol _(for the full protocol and detailed results see: Finnerup et al., 2015 [@Finnerup2015]; Appendix 2):_

* Full reports of randomized, controlled, double-blind studies published in peer-reviewed journals between January, 1966, and April, 2013, were identified by searches of PubMed, Medline, the Cochrane Central Register of Controlled Trials, and Embase. An additional search up to Jan 31, 2014, retrieved papers from PubMed. Additional papers were identified from published reviews and the reference lists of selected papers.

* To identify unpublished trials, studies reporting results were searched in all primary registries in the WHO Registry Network and in registries approved by the International Committee of Medical Journal Editors in April, 2013. Only ClinicalTrials.gov had relevant data. An additional search up to Jan 31, 2014, retrieved studies the ClinicalTrials.gov website. Data from a search in May, 2009, of the Pharmaceutical Research and Manufacturers of America (PhRMA) clinical study results website were also included.

* For the purposes of this application a supplementary search of PubMed was conducted on February 26, 2016. Search terms included: _[drug name]_ pain (randomised or randomized); neuropathic pain and (randomised or randomized); neuralgia and (randomised or randomized); neuropathy pain and (randomised or randomized); not neuropathic. Figure \ref{flow_chart} shows the combined flow chart for study selection from the original search and the update.

\begin{figure}
\centering
\includegraphics[scale = 0.5]{./images/search-flow}
\captionsetup{justification=justified}
\caption{Flow chart of study selection. Updated from Finnerup et al., 2015 [3] in 26 February 2016.}
\label{flow_chart}
\end{figure}

The target population was patients of any age with neuropathic pain according to the International Association for the Study of Pain definition (i.e., pain caused by a lesion or disease of the somatosensory nervous system) [@Jensen2011]: [^3]

[^3]: Post-herpetic neuralgia, diabetic and non-diabetic painful polyneuropathy, post-amputation pain, post-traumatic or post-surgical neuropathic pain including plexus avulsion and complex regional pain syndrome type II (which was generally subsumed into post-traumatic or post-surgical neuropathic pain), central post-stroke pain, spinal cord injury pain, and multiple-sclerosis-associated pain. Neuropathic pain pertaining to different causes was also included. Neuropathic pain associated with nociceptive components (e.g., neuropathic cancer-related pain and radiculopathy) was included if the primary outcome of the study was related to neuropathic pain. Disorders such as complex regional pain syndrome type I, low-back pain without radicular pain, fibromyalgia, and atypical facial pain were not included because they do not meet the current definition of neuropathic pain. Trigeminal neuralgia was assessed separately because the response to drug treatment is generally distinct from other neuropathic pains.

The interventions considered were systemic or topical treatments (oral, sublingual, oropharyngeal, intranasal, topical, subcutaneous, intradermal, and smoking) with at least 3 weeks of treatment. Single-administration treatments with long-term efficacy (high-concentration capsaicin 8% patches, botulinum toxin) were included if there was a minimum follow-up of 3 weeks. Studies in which intramuscular, intravenous, or neuroaxial routes of administration were used and those of pre-emptive analgesia were excluded.

We included randomized, double-blind, placebo controlled studies with parallel group or crossover study designs that had at least ten patients per group. We separately summarised enriched-enrolment, randomized withdrawal trials. We excluded studies published only as abstracts and included double-blind, active comparator trials of drugs generally proposed as first-line or second line treatments. The study outcome (positive or negative) was based on the effect on the primary outcome measure (i.e, neuropathic pain intensity). We excluded studies in which the primary outcome included a composite score of pain and paraesthesia or paraesthesia only.

Studies were assessed for methodological quality by using the five-point Oxford Quality Scale [@Jadad1996]. A minimum score of 2 of 5 (randomized and double-blind study) was required for inclusion [@Jadad1996]. We also assessed the serious risk of bias relating to absence of allocation concealment, incomplete accounting of outcome events, selective outcome reporting, stopping early for benefit, use of invalidated outcome measures, and carry-over effects in crossover trials. We followed the 23-item Appraisal of Guidelines for Research and Evaluation (AGREE II) for developing and reporting recommendations [@Brouwers2010].

Number needed to treat (NNT) for 50% pain intensity reduction (or 30% pain reduction or at least moderate pain relief), calculated using the fixed-effects Mantel-Haenszel method, was the primary effect measure. NNT and NNH were calculated as the reciprocal values. Susceptibility to risk of publication bias was assessed by funnel plots [@Light1984], Egger’s regression [@Egger1997], and Duval and Tweedie’s non-parametric trim-and-fill approach [@Duval2000]. Heterogeneity in trials was presented as a L'Abb$\acute{e}$ plot [@LAbbe1987] and as the I^2 statistic, and heterogeneity, particularly that which was not easily explained by differences in drug dose, diagnosis, and size of placebo response, was included in the GRADE recommendation.

### Evidence summary and reporting

The GRADE classification system was used to summarise the evidence and formulate clinical guidelines [@Guyatt2008; @Balshem2011] with final quality of evidence rated as strong or weak from the summary of available data (appraisal of quality, outcome measures, summary of results).

A total of 229 reports, across a number of agents, were included in the published meta-analysis [@Finnerup2015] [^4]. One hundred and twenty-seven (55%) of 229 trials were in patients with diabetic painful polyneuropathy or post-herpetic neuralgia. NNT could be calculated in 176 (77%) of 229 published placebo-controlled trials.

[^4]: Tricyclic antidepressants (TCAs), serotonin-noradrenaline re-uptake inhibitor antidepressants (SNRIs), other antidepressants, pregabalin, gabapentin or gabapentin extended release and enacarbil, other anti-epileptics, tramadol, opioids, cannabinoids, lidocaine 5% patch, capsaicin high concentration patch and cream, botulinum toxin A, NMDA antagonists, mexiletine, miscellaneous topical treatments, newer systemic drugs, and combination therapies.

The mean Oxford Quality Scale (Jadad) score was 4.1 (SD: 0.87, range: 2 to 5). Funnel plots and Egger regression identified asymmetry. Computing theoretical missing studies using the _'trim-and-fill'_ method suggested about a 10% overstatement of treatment effects across all medicines assessed in the meta-analysis [Figure \ref{bias_1}; 34 theoretical missing studies, which adjusted the effect size from an odds ratio of 1.8 (95% CI 1.7 to 1.9), to 1.6 (95% CI: 1.5 to 1.7)]. Susceptibility to bias analysis of individual medicines/medicine classes confirmed that publication bias was unlikely to be a major confound of this evidence (Figure \ref{bias_2}).

\begin{figure}[ht]
\centering
\includegraphics[scale = 0.4]{./images/bias-1}
\captionsetup{justification=justified}
\caption{Funnel plot showing the precision (inverse of standard error) against the effect size (natural log of the odds ratio, LnOR). Blue circles are individual studies. Missing studies imputed by trim and fill are shown in red. The blue vertical line indicates the uncorrected estimate of the effect size, while the red vertical line indicates the possible summary if the theoretical missing studies included. Adapted from: Finnerup et al., 2015 [3].}
\label{bias_1}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale = 0.54, angle = -90]{./images/bias-2}
\captionsetup{justification=justified}
\caption{Data are number, unless otherwise indicated. $\ast$ Number of comparisons with placebo in published trials and unpublished trials included in the meta-analysis; results from registries were included if they reported numbers of responders. $\dag$ Total number of patients treated with active treatment and placebo; patients were counted twice if the study had a crossover design. $\ddag$ Number of patients needed to be treated in a new study showing no effect to make the number needed to treat (NNT) greater than 11, which is the cut-off for clinical relevance; susceptibility to publication bias implies that a new study with fewer than 400 participants with no effect might increase the NNT to greater than 11. $\mathbb{S}$ Including gabapentin extended release and enacarbil. Adapted from: Finnerup et al., 2015 [3].}
\label{bias_2}
\end{figure}

Using the GRADE process we identified that tricyclic antidepressants (TCAs; mainly amitriptyline) [^5], serotonin–adrenaline re-uptake inhibitors (SNRIs; mainly duloxetine) [^6], pregabalin [^7] and gabapentin could be considered to be first-line drugs (Figure \ref{grade}, and Figure \ref{nnt}). Amitriptyline, a TCA, already features strongly on the WHO Model List of Essential Medicines, and shares its analgesic mechanism of action with other TCAs and SNRIs. Accordingly, all TCAs and SNRIs are contraindicated for use with each other, and this contraindication precludes combination therapy with these medications should patients not respond adequately to monotherapy. Because of the incompatibility of these first-line drug classes, the evidence-base for the use of TCAs and SNRIs is not evaluated further in this section of the application [^15]. Instead, we provide updated information (based on our supplementary search in February 2016) only on the efficacy of the $\alpha 2 \delta$ calcium channel ligands gabapentin and pregabalin. This class of drugs is not contraindicated for use with TCAs or SNRIs, and so may be used alone or in combination therapy with the other two first-line classes of medication. 

[^15]: The supplementary literature search in 2016 identified one new  report on amitriptyline: Dinat et al., 2015 [@Dinat2015]. Dinat and colleagues compared amitriptyline and placebo in HIV-associated sensory neuropathy, and the outcome, which was associated with high placebo responses, was negative for amitriptyline.

\begin{figure}[ht]
\includegraphics[scale = 0.54, angle = -90]{./images/grade-finnerup-2015}
\captionsetup{justification=justified}
\caption{Summary of the GRADE recommendations by Finnerup et al., 2015 [3] for first-line medications for managing neuropathic pain.}
\label{grade}
\end{figure}

[^5]: In 18 placebo-controlled trials [20 comparisons with placebo, of which seven comparisons had active placebos; 12 trials assessed amitriptyline (25–150mg/day)], 16 comparisons were positive. The final quality of evidence was moderate (Appendix 2). There was no evidence of a dose-response effect. Combined NNT for 15 studies was 3.6 (95% CI: 3.0 to 4.4).

[^6]: 14 studies of serotonin-noradrenaline re-uptake inhibitors with available results: nine with duloxetine (20–120 mg, seven positive), four with venlafaxine (doses 150–225 mg/day, two positive, and two negative with low doses), one with venlafaxine (negative; Appendix 2). The final quality of evidence was high. Combined NNT was 6.4 (95% CI: 5.2 to 8.4).

[^7]: 18 of 25 placebo-controlled randomized trials of pregabalin (150–600mg/day) were positive, with high final quality of evidence (Appendix 2). There was a dose response gradient (higher response with 600mg daily than with 300mg daily; data not shown). Combined NNT was 7.7 (95% CI: 6.5 to 9.4). The combined NNT is 8.8 (95% CI: 7.5 to 10.8) when the 5 new studies identified in the 2016 search are included.  

```{r nnt, eval = FALSE}

############################################################
#                                                          #
#    Number need to treat figure: See nnt.R for details    #
#                                                          #
############################################################

```

\begin{figure}[ht]
\includegraphics[scale = 0.45]{./images/nnt}
\captionsetup{justification=justified}
\caption{Mean (95\% CI) numbers needed to treat (NNT) for first-line medications recommended by Finnerup et al., 2015 [3]. The size of the coloured circles indicate the relative number of individuals randomized in trials for a particular medication. Data from Finnerup et al., 2015 were updated to include two new trials in the tricyclic antidepressant class, and gabapentin extended release and enacarbil were excluded from the gabapentin group.}
\label{nnt}
\end{figure}

#### Updated evidence-base for $\alpha 2 \delta$ calcium channel ligands

**Pregabalin**

Eight new reports were identified in the 2016 supplementary search of which one was an enriched-enrollment trial and five provided dichotomous data for NNT calculation. In a mixed peripheral neuropathy population, Holbech and colleagues [@Holbech2015] showed modest analgesic effects for pregabalin (300mg/day) versus placebo and Liu et al 2015 [@Liu2015] found an effect in PHN. The other studies (Simpson et al 2014 [@Simpson2014], Huffman et al. 2015 [@Huffman2015], Raskin et al. 2016 [@Raskin2016], Chappell et al. 2014 [@Chappell2014], and Ziegler et al. 2015 [@Ziegler2015]) failed to find an effect of pregabalin in painful polyneuropathy due to diabetes or HIV. All the negative studies except the study in HIV neuropathy [@Simpson2014] used a 300 mg daily dose of pregabalin. In total, 32 randomized controlled trials of pregabalin for neuropathic pain were identified after our updated search. Thirty of these studies provided dichotomous data, and the updated combined NNT for pregabalin was 8.8 (95% CI: 7.5 to 10.8). There was a dose response gradient (higher response with 600 mg daily than with 300 mg daily).

**Gabapentin**

No additional studies using gabapentin for neuropathic pain were identified in our supplementary search. In total, our assessment was based on 14 randomized controlled trials of gabapentin (900 to 3600 mg/day; nine positive). The trials were predominantly conducted in patients with post-herpetic neuralgia, painful polyneuropathy (mainly diabetic), spinal cord injury, post-amputation pain, and peripheral nerve injury. Combined NNT was 6.3 (95% CI: 5.0 to 8.3) for gabapentin. There was no evidence of a dose-response effect. 

Our data are largely concordant with a recent Cochrane review that evaluated the evidence for the use of gabapentin in chronic neuropathic pain and fibromyalgia in adults [@Moore2014]. The authors concluded on the basis of second tier evidence that gabapentin was efficacious in post-herpetic neuralgia (NNT 8.0, 95% CI: 6.0 to 12) and painful diabetic neuropathy (NNT 5.9, 95% CI: 4.6 to 8.3). The authors concluded that there were insufficient data in other pain conditions, including fibromyalgia, to reach any reliable conclusion.

## Review of harms and toxicity: summary of evidence on safety.

The information on harms and toxicity was obtained from regulatory documents available from the Food and Drug Administration (FDA) [@NeurontinFDA2015], and European Medicines Agency (EMA) [@NeurontinEMA2013] for Neurontin (gabapentin, Pfizer Inc).

#### Contraindications

Gabapentin is contraindicated in patients who have demonstrated hypersensitivity to the drug or its ingredients.

#### Warnings and precautions

**Drug reaction with eosinophilia and systemic symptoms (DRESS):** Although very rare, DRESS has been associated with use of gabapentin. Some of these reactions have been fatal or life-threatening. If signs or symptoms consistent with DRESS present, the patient should be evaluated immediately, and gabapentin discontinued if an alternative aetiology for the signs or symptoms cannot be established. 

**Anaphylaxis and angioedema:** Gabapentin can cause anaphylaxis and angioedema after the first dose or at any time during treatment. Patients should discontinue using gabapentin and seek immediate medical care should they experience signs or symptoms of anaphylaxis or angioedema. 

**Effects on driving and operating heavy machinery:** Patients taking gabapentin should not drive until they have gained sufficient experience to assess whether the medication impairs their ability to drive. Driving performance studies conducted with a pro-drug of gabapentin (gabapentin enacarbil, extended release) indicate that gabapentin may cause significant driving impairment. Prescribers and patients should be aware that patients’ ability to assess their own driving competence, as well as their ability to assess the degree of somnolence caused by gabapentin may be imperfect. The duration of the driving impairment after starting therapy with gabapentin is unknown. Whether the impairment is related to somnolence or other effects of gabapentin is unknown.

Moreover, because gabapentin causes somnolence and dizziness patients should be advised not to operate complex machinery until they have gained sufficient experience on gabapentin to assess whether it impairs their
ability to perform such tasks. 

**Withdrawal precipitated seizure:** Anti-epileptic drugs should not be abruptly discontinued because of the possibility of increasing seizure frequency in individuals with epilepsy. In randomized controlled trials of gabapentin for epilepsy, the incidence of status epilepticus did not differ significantly between participants receiving gabapentin and those receiving placebo. This warning pertains primarily to the use of gabapentin in the context of managing epilepsy rather than neuropathic pain.

**Suicidal behaviour and ideation:** Anti-epileptic drugs (AEDs), including gabapentin, increase the risk of suicidal thoughts or behaviour in patients taking these drugs for any indication. Patients treated with any AED for any indication should be monitored for the emergence or worsening of depression, suicidal thoughts or behaviour, and/or any unusual changes in mood or behaviour. Pooled analyses of 199 placebo-controlled clinical trials (mono- and adjunctive therapy) of 11 different AEDs showed that patients randomized to one of the AEDs had approximately twice the risk of suicidal thinking or behaviour compared to patients randomized to placebo (adjusted relative risk 1.8, 95% CI: 1.2 to 2.7). 

**Tumorigenic potential:** An oral carcinogenicity study of gabapentin in rats increased the incidence of pancreatic acinar cell. The clinical significance of this finding is unknown. 

**Sudden and unexplained death in patients with epilepsy:**  Eight sudden and unexplained deaths were recorded among a cohort of 2203 epilepsy patients treated with gabapentin, a rate of 0.0038 deaths per patient-year. Although this rate exceeds that expected in a healthy population matched for age and sex, it is within the range of estimates for the incidence of sudden unexplained deaths in patients with epilepsy not receiving
gabapentin.

**Somnolence/sedation and dizziness:** In clinical trials for epilepsy, adults receiving doses of up to 1800mg daily of gabapentin reported somnolence, dizziness, and ataxia at a greater rate compared to those receiving placebo (19% in drug versus 9% in placebo for somnolence; 17% in drug versus 7% in placebo for dizziness; and 13% in drug versus 6% in placebo for ataxia). In these trials somnolence, ataxia and fatigue were common adverse reactions leading to discontinuation of gabapentin in patients older than 12 years of age.

#### Adverse events in trials for neuropathic pain

Our analysis of adverse effects in trials of gabapentin for neuropathic pain was based on the 14 studies included in the meta-analysis by Finnerup and colleagues [@Finnerup2015] as our literature search in February 2016 did not identify additional studies. Of the 14 studies, one study used only a low dose of gabapentin (900mg) and two studies did not provide numbers of drop-outs due to side effect, thus the combined number needed to harm (NNH) was based on 11 studies (Figure \ref{nnh_1}). The NNH was calculated as the number of patients who needed to be treated for one patient to drop out because of adverse effects. The 95% confidence intervals (CIs) of the NNH were calculated as the reciprocal values of the 95% CIs for the absolute risk difference using the normal approximation. The combined NNH for gabapentin was 25.6 (95% CI: 15.3 to 78.6) [@Finnerup2015].

\begin{figure}[ht]
\centering
\includegraphics[scale = 0.7]{./images/combined-nnh}
\captionsetup{justification=justified}
\caption{Numbers needed to harm (NNH; 95\% CI) for trial drop-outs due to adverse effects. Data are derived from 11 trials of gabapentin for neuropathic pain. The bottom red diamond shows the combined effect across the 11 trials. The size of the filled squares indicate the relative number of individuals randomized in each trial.}
\label{nnh_1}
\end{figure}

When examining specific adverse events, dizziness, somnolence (or drowsiness or sedation), and in a few studies peripheral oedema and confusion, had a prevalence $>$ 10% and a higher prevalence than in the placebo group. The NNH for dizziness was 5.1 (95% CI: 4.3 to 6.3) and for somnolence 7.1 (95% CI: 5.7 to 9.4) (Figures \ref{nnh_2} and \ref{nnh_3}).

\begin{figure}[ht]
\centering
\includegraphics[scale = 0.7]{./images/dizziness-nnh}
\captionsetup{justification=justified}
\caption{Numbers needed to harm (NNH; 95\% CI) for dizziness. Data are derived from nine trials of gabapentin for neuropathic pain. The bottom red diamond shows the combined effect across the nine trials (5.1, 95\% CI: 4.3 to 6.3). The size of the filled squares indicate the relative number of individuals randomized in each trial.}
\label{nnh_2}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[scale = 0.7]{./images/somnolence-nnh}
\captionsetup{justification=justified}
\caption{Numbers needed to harm (NNH; 95\% CI) for somnolence. Data are derived from nine trials of gabapentin for neuropathic pain. The bottom red diamond shows the combined effect across the nine trials (7.1, 95\% CI: 5.7 to 9.4). The size of the filled squares indicate the relative number of individuals randomized in each trial.}
\label{nnh_3}
\end{figure}

In a Cochrane review of gabapentin in fibromyalgia and neuropathic pain [@Moore2014], 62% during gabapentin and 50% during placebo experienced at least one adverse event in 17 studies with 4002 participants. The risk ratio for adverse events was 1.25 (95% CI: 1.2 to 1.3), and the NNH was 8.6 (95% CI: 6.8 to 12). Serious adverse events were not more common for gabapentin than for placebo (risk ratio = 1.2, 95% CI: 0.8 to 1.7) [@Moore2014]. The NNH for somnolence, drowsiness, or sedation was 11 (95% CI: 9.4 to 14; 4125 participants), for dizziness 7.6 (95% CI: 6.6 to 8.8; 4125 participants), and for peripheral oedema 21 (95% CI: 16 to 30; 3220 participants). Gabapentin was associated with an increased risk of ataxia or gait disturbance with and NNH of 13 (95% CI: 9 to 24; 544 participants) [@Moore2014].

## Summary of efficacy and safety across first-line medications

Table \ref{summary_1} summarises the benefits and harms of gabapentin based on our systematic review and meta-analysis. For comparison, we have also included the data for other medicines we recommended as first-line [@Finnerup2015]. Based on the balance of the evidence, we recommended gapapentin, pregabalin, TCAs and SNRIs as first-line treatments; the updated literature search in 2016 did not change our recommendation. When making our original recommendations, we stated that there was no evidence for any of the agents having superior efficacy in general, or for specific causes of neuropathic pain. Therefore, our recommendations applied to neuropathic pain in general. However, we also noted the paucity of clinical trials on cancer-related neuropathic pain, and the absence of trials in children.   

\begin{table}[ht]
\centering
\caption{Summary of efficacy and adverse events [10]}
\label{summary_1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lccccc@{}}
\toprule
\multirow{2}{*}{}            & \multirow{2}{*}{\begin{tabular}[c]{@{}c@{}}Number needed to treat\\ \footnotesize{(50\% pain relief)}\end{tabular}} & \multicolumn{4}{c}{Number needed to harm}  \\ \cmidrule(l){3-6} 
                             &                                         & Major$^{\ast}$ & Dizziness & Somnolence & Dry mouth \\ \bottomrule
TCA           & 4.3                                     & 13.4  & 10.3      & 9.5        & 4.8       \\
Gabapentin$^{\dag}$ & 6.3                                     & 24.5  & 5.1       & 7.1        & -         \\
Pregabalin                   & 8.8                                     & 13.9  & -         & -          & -         \\
SNRI             & 6.4                                     & 11.8  & -         & -          & -         \\ \bottomrule
\multicolumn{6}{l}{\footnotesize{\begin{tabular}[c]{@{}l@{}}TCA: Tricyclic antidepressants; SNRI: Serotonin and noradrenaline reuptake inhibitors;\\ $\ast$ : Withdrawal from study as a result of adverse events;\\ $\dag$ : Excluding gabapentin extended release / encarbil\end{tabular}}} \\ 
\end{tabular}%
}
\end{table}

In their guideline on the management of neuropathic pain, NICE generated a heat-map of relative benefits and harms of the medications they assessed [@NICE2013]. Figure \ref{nice} presents a summary of that figure that only includes medications recommended as first-line therapy by NICE [@NICE2013] and others [@Finnerup2015; @SIGN2013; @Attal2010]. 

\begin{figure}[h!]
\centering
\includegraphics[scale = 0.5]{./images/nice-summary}
\captionsetup{justification=justified}
\caption{Graphical table showing the probability that each first-line treatment is the best option for which evidence is available, the worst available option, or any point in between. The probabilities are indicated by intensity of colour (see legend). All outcomes presented on a standardised scale, from best (left) to worst (right). Thus, where the outcome is desirable (e.g., pain relief) the treatments with most intense colour in the left-hand part of the scale are those with the highest estimated probability of achieving that result. Where results are for an undesirable outcome (e.g., nausea) a concentration of colour on the left-hand part of the scale implies a lower probability of the event. Relatively pale colours across a broad spread of the scale are indicative of substantial uncertainty, while an intense concentration of colour at one point on the scale reflects unambiguous results. Adapted from: NICE CG173 [4].}
\label{nice}
\end{figure}

## Summary of available data on comparative cost and cost-effectiveness within the pharmacological class or therapeutic group.

### Comparative costs

Comparative pricing data were obtained from the Management Sciences for Health (MSH) International Drug Price Indicator Guide [@MSH2014]. Tables \ref{cost_unit} to \ref{cost_mdd} report comparative prices of gabapentin and two other medications on the WHO Model Essential Medicines List, amitriptyline and carbamazepine. Amitriptyline was included because it is recommended, along with gabapentin as a first-line pharmacological treatment for neuropathic pain [@Finnerup2015; @NICE2013; @SIGN2013; @Attal2010]. Carbamazepine falls into the same therapeutic class as gabapentin (anticonvulsants), and it is recommended for the treatment of trigeminal neuralgia [@Attal2010][^12]. The data are reported as unit price of the medications (Table \ref{cost_unit}), price when prescribed at the defined daily dose for each medication (Table \ref{cost_ddd}), and price when prescribed at the maximum recommended daily dose of the medications (Table \ref{cost_mdd}).

[^12]: In our recent meta-analysis and GRADE analysis [@Finnerup2015] there was inconclusive evidence for the use of carbamazepine in the management of neuropathic pains outside of trigeminal neuralgia, and thus carbamazepine was _not_ recommended for use in the pharmacological management of neuropathic pain. Even in the case of trigeminal neuralgia, the data supporting the use of carbamazepine is old and of low quality [@Attal2010]. 

Analysis of comparative pricing for gabapentin was limited by the absence of price data from suppliers, and price data was only available from one buyer source each for the 100mg and 400mg doses of gabapentin, and three sources for the 300mg dose. 

```{r cost_unit, results = 'asis'}

############################################################
#                                                          #
#          Comparative unit cost of drugs table:           #
#           See cost_unit.R and cost_unit.RData            #
#                      for details                         #
#                                                          #
############################################################

# Load data
load('./data/cost_unit.RData')

# Make xtable
xtab_cost_unit <- xtable(cost_unit, caption = "Price based on the unit cost of gabapentin \\newline \\textit{(amitriptyline and carbamazepine are shown for comparison)}",
                       label = 'cost_unit',
                       # set column widths. Must set R/L/C 
                       # using package array
                       align = c('l', 'L{28mm}', rep('R{17mm}', 2),
                                 rep('R{20mm}', 3)))

# Gray first 3 rows
addtorow <- list(pos = list(0, 1, 2), 
                 command = rep("\\rowcolor[gray]{.9}", 3))

# Print table
print(xtab_cost_unit,
      add.to.row = addtorow)

# Clean environment
rm(list = ls())

```


```{r cost_ddd, results = 'asis'}

############################################################
#                                                          #
#          Comparative ddd cost of drugs table:            #
#           See cost_ddd.R and cost_ddd.RData              #
#                      for details                         #
#                                                          #
############################################################

# Load data
load('./data/cost_ddd.RData')

# Make xtable
xtab_cost_ddd <- xtable(cost_ddd, caption = "Price based on the defined daily dose (DDD) of gabapentin \\newline \\textit{(amitriptyline and carbamazepine are shown for comparison)}",
                       label = 'cost_ddd',
                       # set column widths. Must set R/L/C 
                       # using package array
                       align = c('l', 'L{28mm}', rep('R{17mm}', 2),
                                 rep('R{20mm}', 3)))

# Gray first 3 rows
addtorow <- list(pos = list(0, 1, 2), 
                 command = rep("\\rowcolor[gray]{.9}", 3))

# Print table
print(xtab_cost_ddd,
      add.to.row = addtorow)

# Clean environment
rm(list = ls())

```


```{r cost_mdd, results = 'asis'}

############################################################
#                                                          #
#          Comparative MDD cost of drugs table:            #
#           See cost_mdd.R and cost_mdd.RData              #
#                      for details                         #
#                                                          #
############################################################

# Load data
load('./data/cost_mdd.RData')

# Make xtable
xtab_cost_mdd <- xtable(cost_mdd, caption = "Price based on the maximum daily dose (MDD) of gabapentin \\newline \\textit{(amitriptyline and carbamazepine are shown for comparison)}",
                       label = 'cost_mdd',
                       # set column widths. Must set R/L/C 
                       # using package array
                       align = c('l', 'L{28mm}', rep('R{17mm}', 2),
                                 rep('R{20mm}', 3)))


# Gray first 3 rows
addtorow <- list(pos = list(0, 1, 2), 
                 command = rep("\\rowcolor[gray]{.9}", 3))

# Print table
print(xtab_cost_mdd,
      add.to.row = addtorow)

# Clean environment
rm(list = ls())

```

### Cost-utility analysis

The National Institute of Health and Care Excellence, UK (NICE), recently completed a cost-utility analysis across treatments typically recommended as first-line for neuropathic pain [@NICE2013]. In brief, their methodology included: [^10]

[^10]: For full details on the methodology, please see NICE CG173 guideline [@NICE2013]: \href{https://www.nice.org.uk/guidance/cg173/evidence/appendix-f-191621345}{Appendix F}.

- A literature search of published cost-utility analyses, which yielded 3353 unique citations, 3340 of which were excluded after review, leaving 13 articles (all for peripheral neuropathic pain) for inclusion in the analysis;

- For a drug to be included in the modelling process, at least one estimate of dichotomous pain relief (30% and/or 50% relief compared with baseline) and data on withdrawal due to adverse effects was required;

- Medicine prices were taken from the National Health Service, UK Electronic Drug Tariff register for March 2013, and health benefit was valued in quality-adjusted life-year (QALY).

- Based on the available trial data, a time horizon of 20 weeks was used in the model. And, to take into account the uncertainty associated with each input parameter, the model was built probabilistically using Bayesian Markov-chain Monte-Carlo sampling.

The results of the cost-utility analysis are summarised in Tables \ref{nice_cost} to \ref{nice_qaly30k}. Gabapentin compared favourably with other medications recommended as first-line in the management of neuropathic pain in terms of cost (Table \ref{nice_cost}), and in terms of the probability that it would be considered the most cost-effective option based on an assumed QALY value of $\textsterling$ 20,000 and $\textsterling$ 30,000 (Tables \ref{nice_qaly20k} and \ref{nice_qaly30k}). 

```{r nice_cost, results = 'asis'}

############################################################
#                                                          #
#         NICE CG173 modelling of 140-day cost of          #
#             of drugs: See nice_cost.R and                #
#              nice_cost.RData for details                 #
#                                                          #
############################################################

# Load data
load('./data/nice_cost.RData')

# Make xtable
xtab_nice_cost <- xtable(nice_cost,
                         caption = "NICE health economic model: daily dosages and prices of drugs [4] \\newline \\textit{(amitriptyline, pregabalin, duloxetine, and venlafaxine are shown for comparison)}",
                         label = 'nice_cost',
                         # set column widths. Must set R/L/C 
                         # using package array
                         align = c('l', 'L{30mm}', 'R{25mm}', 'R{50mm}', 'R{25mm}'))

# Gray second row
addtorow <- list(pos = list(1), 
                 command = "\\rowcolor[gray]{.9}")

# Print table
print(xtab_nice_cost,
      # include this weird function to use maths in the rows/columns
      sanitize.text.function = function(x) x, 
      add.to.row = addtorow)

# Clean environment
rm(list = ls())

```


```{r nice_qaly20k, results = 'asis'}

############################################################
#                                                          #
#        NICE CG173 modelling of 20k cost-utility of       #
#            of drugs: See nice_qaly20k.R and              #
#             nice_qaly20k.RData for details               #
#                                                          #
############################################################

# Load data
load('./data/nice_qaly20k.RData')

# Make xtable
xtab_nice_qaly20k <- xtable(nice_qaly20k,
                            caption = "NICE health economic model: Probabilistic sensitivity analysis \\newline when \\textbf{1 QALY is valued at $\\textsterling$ 20,000} [4] \\newline \\textit{(amitriptyline, pregabalin, duloxetine, and venlafaxine are shown for comparison)}",
                            label = 'nice_qaly20k',
                            # set column widths. Must set R/L/C 
                            # using package array
                            align = c('l', 'L{35mm}', 'R{30mm}', 'R{30mm}', 'R{35mm}'),
                            digits = c(0, 0, 0, 1, 1))

# Gray second row
addtorow <- list(pos = list(1), 
                 command = "\\rowcolor[gray]{.9}")

# Print table
print(xtab_nice_qaly20k,
      # include this weird function to use maths in the rows/columns
      sanitize.text.function = function(x) x, 
      add.to.row = addtorow)

# Clean environment
rm(list = ls())

```


```{r nice_qaly30k, results = 'asis'}

############################################################
#                                                          #
#        NICE CG173 modelling of 30k cost-utility of       #
#            of drugs: See nice_qaly30k.R and              #
#             nice_qaly30k.RData for details               #
#                                                          #
############################################################

# Load data
load('./data/nice_qaly30k.RData')

# Make xtable
xtab_nice_qaly30k <- xtable(nice_qaly30k,
                            caption = "NICE health economic model: Probabilistic sensitivity analysis \\newline when \\textbf{1 QALY is valued at $\\textsterling$ 30,000} [4] \\newline \\textit{(amitriptyline, pregabalin, duloxetine, and venlafaxine are shown for comparison)}",
                            label = 'cost_3',
                            # set column widths. Must set R/L/C 
                            # using package array
                            align = c('l', 'L{35mm}', 'R{30mm}', 'R{30mm}', 'R{35mm}'),
                            digits = c(0, 0, 0, 1, 1))

# Gray second row
addtorow <- list(pos = list(1), 
                 command = "\\rowcolor[gray]{.9}")

# Print table
print(xtab_nice_qaly30k,
      # include this weird function to use maths in the rows/columns
      sanitize.text.function = function(x) x, 
      add.to.row = addtorow,
      table.placement = 'h!')

# Clean environment
rm(list = ls())

```

Based on the outcome of the cost-utility analysis, the NICE Guideline Development Group recommended gabapentin and amitriptyline as initial treatment options for neuropathic pain. 

The results of the NICE cost-utility analysis, combined with similar efficacy and safety profiles for the molecules, informed our decision to apply for inclusion of gabapentin on the Model List, and not pregabalin, the other agent in the $\alpha2\delta$-calcium channel ligand class. Although pregabalin, unlike gabapentin, demonstrates a linear absorption profile and has a universal indication for treatment of neuropathic pain by stringent regulatory bodies, we concluded that, on the balance of the core GRADE indicators of cost, efficacy, and safety gabapentin was the more suitable agent for widespread recommendation at present.

# Regulatory information

## Summary of regulatory status of the medicine.

Gabapentin has regulatory approval as a **prescription only medicine** from the following stringent regulatory bodies: US Federal Drug Administration (FDA), European Medicines Agency (EMA), Australian Therapeutic Goods Administration (TGA), Japanese Pharmaceuticals and Medical Devices Agency (PMDA), and Health Canada (see Table \ref{reg_approval} for registered neuropathic pain indications [^11]). None of these agencies have registered gabapentin as a controlled substance. 

```{r reg_approval, results = 'asis'}

############################################################
#                                                          #
#         Regulatory approval: See reg_approval.R          #
#           and reg_approval.RData for details             #
#                                                          #
############################################################

# Load data 
load('./data/reg_approval.RData')

# Make xtable    
xtab_regulatory <- xtable(reg_approval,
                          caption = "Regulatory approval of gabapentin for neuropathic pain by \\newline major national and regional regulatory bodies ",
                          label = 'regulatory',
                          # set column widths. Must set R/L/C 
                          # using package array
                          align = c('l', 'L{48mm}', 'L{27mm}', 'L{70mm}'))

# Gray second row
addtorow <- list(pos = list(0, 2, 4), 
                 command = rep("\\rowcolor[gray]{.9}", 3))

# Print table
print(xtab_regulatory,
      add.to.row = addtorow,
      table.placement = 'ht')

# Clean environment
rm(list = ls())

```

[^11]: All four regulatory authorities indicate gabapentin as adjunctive therapy in the treatment of partial seizures with and without secondary generalization in adults and children. The PMDA and EMA also indicate gabapentin as mono-therapy for partial seizures with and without secondary generalization in adults and children. 

## Availability of pharmacopoeial standards (British Pharmacopoeia, International Pharmacopoeia, United States Pharmacopoeia, European Pharmacopeia).

Pharmacopoeial standards for gabapentin are included in the:

- United States Pharmacopoeia (USP)

- European Pharmacopoeia (PhEur)

\newpage

# References
\setlength{\parindent}{-10mm}
\setlength{\leftskip}{20mm}
\setlength{\parskip}{8pt}
\noindent

